---
title: "NCbirths Analysis"
author: "Matt Robinson, Athena Higgins, Ethan Sennett, Peter Nguyen"
date: "11/30/2016"
output: 
  pdf_document:
    toc: true
---

```{r setup, echo=FALSE}
library(ggplot2)
library(Stat2Data)
data(NCbirths)
NCbirths1<-na.omit(NCbirths)

```

\newpage

# 1. Group Analysis
## Project Overview

For our final project, based on the NCbirths dataset, we will explore factors related to childbirth data at a particular hospital. The NCbirths dataset includes the following variables that will be useful in exploring the stories contained in this dataset.

**Variable**   |  **Explanation**
-------------- | -----------------
MomAge         | The mothers age.
Marital        | The marital status of the mother. 1 = married, 2 = not married.
RaceMom        | Mother's race: 1=white, 2=black, 3=American Indian, 4=Chinese, 
&nbsp;         | 5=Japanese, 6=Hawaiian, 7=Filipino, 8=Other Asian or Pacific Islander
Gained         | The amount of weight gained during pregnancy (in pounds).
BirthWeightOz	 | The birth weight of the child (in ounces).
Weeks          | The number of completed weeks of gestation.
Plural         | Multiple birth indicator. 1=single birth, 2=twins, 3=triplets.



Examining relationships among these factors provides insight into the complexity of pregnancy for women and their infants, and the plethora of variables which influence childbirth. Some interesting relationships and interactions that we will explore in this report include:

1. The relationship between the mothers marital status, race and age.
2. The relationship between the mothers marital status, weight gained during pregnancy and age.
3. The relationship between the mothers marital status, the childs weight at birth and number of weeks of gestation.
4. The relationship between the mothers marital status, the childs weight at birth and whether or not the child was a twin or triplet.

Analyzing the data, through lenses shaped by these factors, allows us to tell intriguing social and biological stories about women’s pregnancy experiences. All of these analyses will incorporate marital status as a categorical factor to provide a comprehensive analysis of marital influence on pregnancy and childbirth.

We will use our analysis to prove or disprove the following hypotheses regarding marital status and pregnancy:

1. Hypothesis 1 -  Births involving a single mother will exhibit a higher weight gained during pregnancy, a lower average of weeks of gestation and a lower average child weight at birth.

2. Hypothesis / Exploration 2- Interpreting data related to marital status, race, and age will allow for an intriguing discussion of social factors related to childbirth, and insight into the influence of these factors on pregnancy. Marital status and race are not biological factors, and thus should have no physical influence on childbirth. When different patterns emerge in the childbirth data, with respect to these variables, trends could indicate the influence of social pressures and climate on childbirth. 

3. Hypothesis 3 - Births involving a married mother where the child will have a higher average weight, and will be a twin. 



## Missing Data
```{r, echo=FALSE,eval=FALSE}
# Code for Missing Data
head(NCbirths[!complete.cases(NCbirths),])
```

Using R, we were able to find 41 observations where at least 1 variable had missing data. We decided to discard all 41 observations, since it wouldn't be taking too much away from the whole population in the dataset. Therefore, all of our visualizations and analyses use this cleaned dataset (NCbirths1).


## Visualization of Data

The below plots provide some basic visualiazations of the data collected. The sample collected for this dataset provides ample sample points for the various values the variables can take on.


```{r, fig.height=2, fig.width= 3.6, warning=FALSE, echo=FALSE}
ggplot(NCbirths1, aes(MomAge, Weeks, color=factor(MomRace))) + geom_point(size=.5) + labs(color="Race")

qplot(Gained, data = NCbirths1, geom = "density", color = factor(Marital)) + ylab("") + labs(color="Marital")
```

```{r fig.height = 3, fig.width = 3.6,echo=FALSE}
ggplot(NCbirths1, aes(MomAge, Marital, color = RaceMom)) + geom_point() 

ggplot(NCbirths1, aes(MomAge, Gained, color=factor(Marital))) + geom_point() 
```

