---
title: "NCbirths Analysis"
author: "Matt Robinson, Athena Higgins, Ethan Sennett, Peter Nguyen"
date: "12/7/2016"
output:
  pdf_document:
    toc: yes
---

```{r setup, echo=FALSE}
library(ggplot2)
library(Stat2Data)
library(splines)
library(knitr)
data(NCbirths)
NCbirths1<-na.omit(NCbirths)
NCbirths1$marital_cat <- factor(NCbirths1$Marital, levels=c(1,2), labels=c("married", "not married"))

```

\newpage

# 1. Group Analysis
## Project Overview

For our final project, based on the NCbirths dataset, we will explore factors related to childbirth data at a particular hospital. The NCbirths dataset includes the following variables, which will be useful in exploring the stories contained within this dataset.

**Variable**   |  **Explanation**
-------------- | -----------------
MomAge         | The mothers age.
Marital        | The marital status of the mother. 1 = married, 2 = not married.
RaceMom        | Mother's race: 1=white, 2=black, 3=American Indian, 4=Chinese, 
&nbsp;         | 5=Japanese, 6=Hawaiian, 7=Filipino, 8=Other Asian or Pacific Islander
Gained         | The amount of weight gained during pregnancy (in pounds).
BirthWeightOz	 | The birth weight of the child (in ounces).
Weeks          | The number of completed weeks of gestation.
Plural         | Multiple birth indicator. 1=single birth, 2=twins, 3=triplets.


Examining relationships among these factors provides insight into the complexity of pregnancy for women and their infants, and the plethora of variables which influence childbirth. Some interesting relationships and interactions that we will explore in this report include:

1. The relationship between the mothers marital status, race and age.
2. The relationship between the mothers marital status, weight gained during pregnancy and age.
3. The relationship between the mothers marital status, the childs weight at birth and number of weeks of gestation.
4. The relationship between the mothers marital status, the childs weight at birth and whether or not the child was a twin or triplet.

Analyzing the data, through lenses shaped by these factors, allows us to tell intriguing social and biological stories about women's individual pregnancy experiences. Each individual analysis of the dataset incorporates marital status as a categorical factor, in order to provide some consistency throughout analysis, and to allow for comprehensive overview of marital influence on pregnancy and childbirth.

We will use our analysis to prove or disprove the following hypotheses regarding marital status and pregnancy:

1. Hypothesis 1 -  Births involving a single mother will exhibit a higher weight gained during pregnancy, a lower average of weeks of gestation and a lower average child weight at birth.

2. Hypothesis / Exploration 2- Interpreting data related to marital status, race, and age will allow for an intriguing discussion of social factors related to childbirth, and insight into the influence of these factors on pregnancy. Marital status and race are not biological factors, and thus should have no physical influence on childbirth. When different patterns emerge in the childbirth data, with respect to these variables, trends could indicate the influence of social pressures and climate on childbirth. 

3. Hypothesis 3 - Births involving a married mother where the child will have a higher average birth weight, and will be a twin. On average, a married family will have a higher income than someone who is not married. When this is taken into account, food intake for the family is affected so by being married, the mother will be able to eat more and in turn the children will have a higher birth weight.



## Missing Data

```{r, echo=FALSE,eval=FALSE}
# Code for Missing Data
head(NCbirths[!complete.cases(NCbirths),])
```

Using R, we were able to find 41 observations where at least 1 variable had missing data. We decided to discard all 41 observations, because eliminating these entries would not remove an extensive amount of data from the entire collection of data. We also examined the individual entries missing data, and concluded that there was not common trend among participants for whom information was missing. (If there had been a trend- for example, if all Japanese mothers neglected to provide their marital status, then removing data could have posed a problem when analyzing the data.) All of the visualizations and analysis conducted for this paper utalize the "clean" / "complete" dataset. 


## Visualization of Data

The plots below provide some basic visualiazations of the data collected, to gain some familiarity with the dataset. 


```{r, fig.height=3, fig.width= 3.6, warning=FALSE, echo=FALSE}
ggplot(NCbirths1, aes(MomAge, Weeks, color=factor(MomRace))) + geom_point(size=.7) + geom_smooth(method="lm", se=FALSE)+ labs(color="Race")

qplot(Gained, data = NCbirths1, geom = "density", color = factor(Marital)) + ylab("") + labs(color="Marital")
```

```{r fig.height = 4, fig.width = 3.6, echo=FALSE}
ggplot(NCbirths1, aes(MomAge, Marital, color = RaceMom)) + geom_point() + labs(color = "Race")

ggplot(NCbirths1, aes(MomAge, Gained, color=factor(Marital))) + geom_point(size=0.7) + labs(color="Marital")
```


\newpage

#2. Individual Analyses
## Variables to Consider: Marital Status, RaceMom, MomAge (Athena Higgins)

Note:
RaceMom	1=white, 2=black, 3=American Indian, 4=Chinese, 5=Japanese, 6=Hawaiian, 7=Filipino, or 8=Other 

```{r, include = FALSE}
NCbirths_white = subset(NCbirths1, RaceMom == 1)
NCbirths_black = subset(NCbirths1, RaceMom == 2)
NCbirths_american_indian = subset(NCbirths1, RaceMom == 3)
NCbirths_chinese = subset(NCbirths1, RaceMom == 4)
NCbirths_japanese = subset(NCbirths1, RaceMom == 5)
NCbirths_hawaiian = subset(NCbirths1, RaceMom == 6)
NCbirths_filipino = subset(NCbirths1, RaceMom == 7)
NCbirths_other = subset(NCbirths1, RaceMom == 8)

MomAge_cat <- cut(NCbirths$MomAge, breaks = c(0,10,20,30,40,50), labels = c("0-10", "10-20", "20-30", "30-40", "40-50"))
```

```{r, fig.height=2, fig.width= 3.7, warning=FALSE, echo=FALSE}
ggplot(NCbirths1, aes(RaceMom), color = factor(cyl)) + geom_histogram() + ggtitle("Data Collected for Racial Groups")

ggplot(NCbirths1, aes(RaceMom, MomAge, color = factor(Marital))) + geom_point() + ggtitle("Age, Race and Marital Status")

#plot(MomAge_cat) + geom_title("Distribution of Mother's Age")
```
These graphics illustrate the pitfalls of conducting a convenience sample. The groups represented clearly neglect certain subsets of the population (for example, hawaiian individuals are not represented in the dataset). It is important to keep this oversimplification in mind, when interpreting any statistics computed 

This distribution takes on new meaning when factored by marital status and reexamined. Applying the T-test, we can compare the mean MomAge value computed for different subsets within the entire population. 

```{r, include = FALSE, eval = FALSE, tidy = TRUE}

#Statistics on age, factored by Race
age_white <- summary(subset(NCbirths, RaceMom == 1, select = MomAge))
age_black <- summary(subset(NCbirths, RaceMom == 2, select = MomAge))
age_american_indian <- summary(subset(NCbirths, RaceMom == 3, select = MomAge))
age_chinese <- summary(subset(NCbirths, RaceMom == 4, select = MomAge))
age_japanese <- summary(subset(NCbirths, RaceMom == 5, select = MomAge))
age_hawaiian <- summary(subset(NCbirths, RaceMom == 6, select = MomAge))
age_filipino <- summary(subset(NCbirths, RaceMom == 7, select = MomAge))
age_other <- summary(subset(NCbirths1, RaceMom == 8, select = MomAge))
```

The following are mean age values for new mothers, categorized by their race. 
```{r, include = FALSE, tidy = TRUE, eval = FALSE}
#Additional Comparison of Means
population_mean =  summary(subset(NCbirths, select = MomAge))
white_mean = age_white[4]
black_mean = age_black[4]
american_indian_mean = age_american_indian[4]
chinese_mean = age_chinese[4]
japanese_mean = age_japanese[4]
hawaiian_mean = age_hawaiian[4]
filipino_mean = age_filipino[4]
other_mean = age_other[4]
```
**Race**        |**MomAge**           | **Race**        |  **MomAge**      | **Race**      |**MomAge**   | **Race** |**MomAge**
--------------  | ----------          |-----------      |------------------|---------------|-------------|----------|----------
population      | 26.76               |chinese          | 31.0             | black         | 24.53       |filipino  |    37  
white           | 27.92               |japanese         | 24.75            |american indian| 24.09       |other     | 28.86 

The T-Test can also be applied to evaluate the statistical significance of variation observed in means, for diferent subsets of the populations. These computations allow for extraction of a pvalue (pvalue gives the probability of obtaining a result equal to or more extreme than the outcome observed, under the the assumption that the null hypothesis is true.) The T-Test null hypothesis states that the difference in mean values is not statistically significant. However, if the pvalue is very small (less than alpha), we can reject the null hypothesis, and conclude that the difference in mean values is statistically significant. 
```{r, include = FALSE, tidy = TRUE, eval = FALSE}

ttest_white_population <- t.test(subset(NCbirths, RaceMom == 1, select = MomAge), subset(NCbirths, select = MomAge))
t_white = ttest_white_population[3]

ttest_black_population <- t.test(subset(NCbirths, RaceMom == 2, select = MomAge), subset(NCbirths, select = MomAge))
t_black = ttest_black_population[3]

ttest_american_indian_population <- t.test(subset(NCbirths, RaceMom == 3, select = MomAge), subset(NCbirths, select = MomAge))
t_american_indian = ttest_american_indian_population[3]

ttest_chinese_population <- t.test(subset(NCbirths, RaceMom == 4, select = MomAge), subset(NCbirths, select = MomAge))
t_chinese = ttest_chinese_population[3]

ttest_japanese_population <- t.test(subset(NCbirths, RaceMom == 5, select = MomAge), subset(NCbirths, select = MomAge))
t_japanese = ttest_japanese_population[3]

#ttest_hawaiian_population <- t.test(subset(NCbirths, RaceMom == 6, select = MomAge), subset(NCbirths, select = MomAge))
#t_hawaiian = ttest_hawaiian_population

#ttest_filipino_population <- t.test(subset(NCbirths, RaceMom == 7, select = MomAge), subset(NCbirths, select = MomAge))
#t_filipino = ttest_filipino_population

ttest_other_population <- t.test(subset(NCbirths, RaceMom == 8, select = MomAge), subset(NCbirths, select = MomAge))
t_other = ttest_other_population[3]
```
Evaluating the difference in mean MomAge between specific racial groups and the entire population

**Race    **    |  **Pvalue**      | **Significance**     |                |                  |
----------------| ---------------- | ---------------------|----------------|----------------  |---------------------
white           | 4.730464e-06     | Reject the null      |japanese        | 1.335714e-05     | Reject the null      |
black           | 1.486671e-09     | Reject the null      |hawaiian     	 | NaN              |
american indian | 0.03067128       | Depending on alpha   |filipino        | NaN              |  
chinese         | NaN              |                      |other           | 0.1550142        | Fail to reject
white vs. black | 4.78133e-18      | Reject the null

```{r, include = FALSE, tidy = TRUE, eval = FALSE}
#T Test, Comparing MomaAge White and Black MomAge
ttest_white_black <- t.test(subset(NCbirths, RaceMom == 1, select = MomAge), subset(NCbirths, RaceMom == 2, select = MomAge))
t_white_black = ttest_white_black[3]

```
The T-Test can also be used to determine if there is a statistically significant difference in MomAge between racial groups. Comparing white and black mothers, the p-value computed (4.78133e-18) prompts us to reject the null hypothesis, and conclude that there is a statistically significant difference in average age of mother giving birth. 

Interesting observations follow from executing T-Tests to compare MomAge for married and non-married women, for different subsets of the population (for example- mean MomAge for white married women vs. white non-married women). 

Comparing MomAge between married and non-married women (factored by race)
```{r, include = FALSE, tidy = TRUE, eval = FALSE}
#T Test, Entire Population- Married vs. Not-Married
ttest_population_married <- t.test(subset(NCbirths, Marital == 1, select = MomAge), subset(NCbirths, Marital == 2, select = MomAge))
t_population_agemarried = ttest_population_married[3]

#T Test, White- Married vs. Not-Married
ttest_white_married <- t.test(subset(NCbirths, RaceMom == 1 & Marital == 1, select = MomAge), subset(NCbirths, RaceMom == 1 & Marital == 2, select = MomAge))
t_white_agemarried = ttest_white_married[3]

#T Test, Black- Married vs. Not-Married
ttest_black_married <- t.test(subset(NCbirths, RaceMom == 2 & Marital == 1, select = MomAge), subset(NCbirths, RaceMom == 2 & Marital == 2, select = MomAge))
t_black_agemarried = ttest_black_married[3]

#T Test, American indian- Married vs. Not-Married
ttest_american_indian_married <- t.test(subset(NCbirths, RaceMom == 3 & Marital == 1, select = MomAge), subset(NCbirths, RaceMom == 3 & Marital == 2, select = MomAge))
t_american_indian_agemarried = ttest_american_indian_married[3]

#T Test, chinese- Married vs. Not-Married
#NOT ENOUGH DATA
#ttest_chinese_married <- t.test(subset(NCbirths, RaceMom == 4 & Marital == 1, select = MomAge), subset(NCbirths, RaceMom == 4 & Marital == 2, select = MomAge))
#t_chinese_agemarried = ttest_chinese_married[3]

#T Test, japanese- Married vs. Not-Married
ttest_japanese_married <- t.test(subset(NCbirths, RaceMom == 5 & Marital == 1, select = MomAge), subset(NCbirths, RaceMom == 5 & Marital == 2, select = MomAge))
t_japanese_agemarried = ttest_japanese_married[3]

#T Test, hawaiian- Married vs. Not-Married
#NOT ENOUGH OBSERVATIONS
#ttest_hawaiian_married <- t.test(subset(NCbirths, RaceMom == 6 & Marital == 1, select = MomAge), subset(NCbirths, RaceMom == 6 & Marital == 2, select = MomAge))
#t_hawaiian_agemarried = ttest_hawaiian_married[3]

#T Test, filipino- Married vs. Not-Married
#NOT ENOUGH OBSERVATIONS
#ttest_filipino_married <- t.test(subset(NCbirths, RaceMom == 7 & Marital == 1, select = MomAge), subset(NCbirths, RaceMom == 7 & Marital == 2, select = MomAge))
#t_filipino_agemarried = ttest_filipino_married[3]

#T Test, other- Married vs. Not-Married
ttest_other_married <- t.test(subset(NCbirths, RaceMom == 8 & Marital == 1, select = MomAge), subset(NCbirths, RaceMom == 8 & Marital == 2, select = MomAge))
t_other_agemarried = ttest_other_married[3]
```
**Race    **    |  **Pvalue**      | **Significance**
----------------| ---------------- | ------------------------------------------------------ 
white           | 4.003403e-30     | Reject the null- statistically significant difference
black           | 4.793121e-18     | Reject the null- statistically significant difference
american indian | 3.040416e-05     | Reject the null- statistically significant difference
japanese        | 0.05769045       | Fail to Reject
other           | 0.0001803798     | Reject the null- statistically significant difference
Population      | 2.929577e-72     |

(insufficient data to compute p-values for chinese, hawaiian, and filipino subsets)
The pvalues computed, and the conclusions drawn from these pvalues, allow for unique insight into the dataset. For example, the Japanese subset is the only population for which the age of married vs. not-married mothers is not statistically significant. The power of the T-Test lies in its ability to make sense of large datasets, and to determine which differences are meaningful (to be distinguished from natural variation wgucg exusts throuhout the population). 

Another powerful story emerges when we build a regression model to predict MomAge from MomRace and Marital Status. This regression is based on the concept of variable interaction (which states that MomRace and Marital-status interact in unique ways, to influence predictions made of MomAge). The regression will result in B values for each possible combination (example- white married, white not married, black married, black not married etc), which indicates the difference in mean mom-age for that specific group, compared to the population at large. The regression models (MomAge - Population_mean_age), the difference between MomAge for a specific subset, and the overall average population age. This regression does not require an intercept (because it is categorical, not continuous data), hence the -1 term. 
```{r, eval = FALSE, echo = FALSE, tidy = TRUE}

#based on computation above, population_mean = 26.76
#PROBLEM WITH THIS- there is no order / meaning to organization of races (white, black, chinese...)
Marital_factor <- factor(NCbirths1$Marital, levels = c(1,2), labels = c("married", "not married"))

RaceMom_factor <- factor(NCbirths1$RaceMom, levels = c(1,2,3,4,5,6,7,8), labels = c("White", "Black", "American Indian", "Chinese", "Japanese", "Hawaiian", "Filipino", "Other"))

MomAgeFit_RaceMom_Marital <- lm((MomAge - 26.76) ~ (RaceMom_factor * Marital_factor) - 1, data = NCbirths1)
MomAgeFit_RaceMom <- lm((MomAge - 26.76) ~ RaceMom_factor - 1, data = NCbirths1) 
MomAgeFit_Marital <- lm((MomAge - 26.76) ~ Marital_factor - 1, data = NCbirths1)

plot(RaceMom_factor, Marital_factor)

MomAgeFit_RaceMom_Marital
#MomAgeFit_RaceMom
#MomAgeFit_Marital

```
```{r, eval = FALSE, tidy = TRUE}
MomAgeFit_RaceMom_Marital <- lm((MomAge - 26.76) ~ (RaceMom_factor * Marital_factor) - 1, data = NCbirths1)
```
Coefficients Observed 

**Subset- married**         |  **Coefficient** | **Subset- not married**      | **Coefficient**
----------------------------| ---------------- | -----------------------------|-----------------
White Married               |  2.4118          |  White Not Married           | -6.2118      
Black Married               |  1.7036          |  Black Not Married           |  0.3761          
American Indian Married     |  2.2400          |  American Indian Not Married |	-1.5025     
Japanese Married            | -1.1067          |  Japanese Not Married        |  4.4156 
Filipino Married            |  10.2400         |  Filipino Not Married        |  NA 
Chinese Married             |  4.2400          |  Chinese Not Married         |  NA
Hawaiian Married            |  NA              |  Hawaiian Not Married        |  NA
Other Married               |  4.5341          |  Other Not Married           | -4.4823 

Each coefficient observed reveals the difference in MomAge between the particular subset being considered, and the average MomAge for the entire population sampled (26.76). For example, the coefficient computed for white married individuals (2.4118) indicates that white married individuals tend to have children 2.4118 years after the average individual, based on the entire population sampled. On average, White non-married individuals are predicted to have children 6.2118 years before (younger than) the average person, based on coefficients produced from this regression. Each of the coefficients provide insight into the interaction that occurs between these two sociological variables (race and marital status), and the influence this interaction has when when predicting MomAge.
\newpage


## Marital Status, Child Weight &  Weeks of Gestation (Matt Robinson)

In this section, we will explore the relationship between the weight of the child at birth and the number of weeks of gestation. We will also use marital status as a categorical variable to explore whether or not there exists a relationship between marital status and child weight.

```{r, fig.height=2, fig.width= 3.7, warning=FALSE, echo=FALSE}
qplot(Weeks, data = NCbirths1, geom = "density", color = factor(marital_cat)) + ylab("") + labs(color="Marital")

qplot(BirthWeightOz, data = NCbirths1, geom = "density", color = factor(marital_cat)) + ylab("") + labs(color="Marital")
```

```{r, fig.height=3, fig.width=6.5, echo=FALSE}
ggplot(NCbirths1, aes(Weeks, BirthWeightOz, color=factor(marital_cat))) + geom_point() + geom_smooth(method='lm', se=FALSE) + labs(color="Marital")
```

The above plots confirm that the data we will be looking at is normally distributed. Some key insight we gain from the density plot on the right and the scatterplot on the bottom is that there exists a larger mean birth weight in births involving a married mother compared to births involving a non-married mother. By fitting a linear model, we can get a more tangible value for this relationship.

```{r, include=FALSE}
a = lm(BirthWeightOz ~ Weeks + Marital, data=NCbirths1)
b = lm(BirthWeightOz ~ ns(Weeks, df=4) + Marital, data=NCbirths1)
wks <- seq(22,45,1)
newData <- data.frame(Weeks=c(wks, wks), Marital=c(rep(1, times=length(wks)), rep(2, times=length(wks))))
newData$pred <- predict(b, newdata = newData, type="response")
newData$marital_cat <- factor(newData$Marital, levels=c(1,2), labels=c("married", "not married"))
```

```{r, echo=FALSE}
kable(summary(a)$coef, digits=2)
```

From this linear model fit, we can see that for an increase of 1 week of gestation, we would expect a newborn to weigh, on average, 4.8 Oz more. Additionally, the marital coefficient tells us that, on average, non-married mothers birth children that weigh 6 Oz less than married mothers. An important statistic to pay attention to in a linear regression model is the adjusted R-squared value. This linear model was fit with an $R^2_{adj}$ value of 0.3581.

###Evaluation of Linear Regression Model

```{r, fig.height=1.5, fig.width= 3.2, warning=FALSE, echo=FALSE}
qplot(Weeks, a$residuals, data=NCbirths1) + geom_hline(yintercept=0, linetype=2)
ggplot(a, aes(a$residuals)) +
  geom_histogram(binwidth=10)
```
The above plots are useful for checking the conditions for linear regression models. From the right plot, we can see that there appears to be a normal distribution of residuals. The left plot gives us an indication that this relationship may not be linear, there appears to be a slight curve in the residuals plot. This could be due to missing data, as both the residual plot and the scatterplot from the beginning show us that there is a large amount of data collected on births involving 35 or more weeks of gestation, but not much data on lower weeks. If we had more data, the model would be less influenced by high-leverage points on both the lower end and higher end of the spectrum. 


```{r, fig.height=3, fig.width=6, echo=FALSE}
ggplot() + geom_line(data=newData, aes(Weeks, pred, color=factor(marital_cat))) + geom_point(data=NCbirths1, aes(Weeks, BirthWeightOz, color=factor(marital_cat))) + labs(color="Marital")
```

```{r, fig.height=1.5, fig.width= 3.2, warning=FALSE, echo=FALSE}
qplot(Weeks, b$residuals, data=NCbirths1) + geom_hline(yintercept=0, linetype=2)
ggplot(b, aes(b$residuals)) +
  geom_histogram(binwidth=10)
```

We can fit a slightly better model using a natural spline, which fits a curved line to the data. This ends up explaining 8% more of the variability, yielding an $R^2_{adj}$ of 0.4246 when using 4 degrees of freedom. This new model also meets the model diagnostics criteria, and specifically fits the portion of the data with fewer sample points.



\newpage

## Martial Status, Child Weight & Multiple Births (Peter Nguyen)
For this segment, we are looking at the relationship between marital the weight of the child and if there was multiple births. Like with the other segements, we are using the marital status of the mother to see if there is any relationship between them. 

```{r,fig.height = 3, fig.width = 3.6,echo=FALSE}
multi <- factor(NCbirths1$Plural, levels = c(1,2,3), labels = c("single", "twins", "triplets"))
multiCut <- cut(NCbirths1$Plural, breaks = c(0,1,2,3), labels = c("single", "twins", "triplets"))
table(multiCut)
married <- factor(NCbirths1$Marital, levels = c(1,2), labels = c("Married", "Non-Married"))
ggplot(NCbirths1, aes(Plural, BirthWeightOz, color=factor(Marital))) + geom_point() + geom_smooth(method = 'lm',span = 3) + labs(color="Marital")
tapply(NCbirths1$Plural, NCbirths1$Marital, FUN = sum)
```

Looking at the graph from above, we can see that there is a trend with BirthWeightOz that is being affected by Marital status of the mother and multiple births. By having a linear model, we can visualize the trend of the data which happens to be a negative trend with the married mother having a higher average birth weight than a single mother. We also note that being a married mother, the chances of having multiple children during a single birth is higher than a single mother they are more likely to have atleast one child. There could be some explaination for this where the single mother thought that she could not be able to provide for multiple children and chose not to keep them. 

```{r, fig.height = 3, fig.width = 3.6,echo=FALSE}
qplot(factor(Marital), BirthWeightOz, data = NCbirths1, geom = "boxplot")
tapply(NCbirths1$BirthWeightOz, NCbirths1$Marital, FUN = mean)
qplot(factor(Plural), BirthWeightOz, data = NCbirths1, geom = "boxplot")
tapply(NCbirths1$BirthWeightOz, NCbirths1$Plural, FUN = mean)

```
Looking at the box and whisker plots from above, we can see the relationship between the weight of the child at birth and the marital status of the mother. This data was inline with our third hypothesis where being married would have a positive affect on the average weight of the child at birth. Looking at the initial box plot we can see that the average weight of the child is 119.1oz for married mothers and 111.38oz for single mothers. We can also see from the second plot where with each subsequent child, the average weight of the children are much lower. Looking at the average weight of each children for multiple births, we can see a significant drop in average weight from 117.85oz for single children to 76.91oz for twins and 62.5oz for triples. There could be multiple explanations for why this is the case from our data. One could argue that we do not have enough data due to the vast majority of births end in a single child where twins are rare and having triplets are even rarer. One can also believe that with each extra child during the same birth, the mother may not have eaten enough nutrients for each baby.    

###Regression Model
````{r, tidy = TRUE, eval = FALSE}
birthWeight = lm(BirthWeightOz ~ Plural + Marital, data=NCbirths1)
summary(birthWeight)
```
From looking at the summary of the birthWeight, we would expect having multiple births will reduce the BirthWeightOz of the baby if all other variables stay constant. It is also important to note that the fit of the linear model which is given by the r-square value which is .1457.

\newpage

## Marital Status, Age of Mother, and Weight Gained (Ethan Sennett)

```{r,echo=FALSE}
data(NCbirths)
NCbirths2<-NCbirths1[which(NCbirths1$MomAge <= 27 & NCbirths1$Marital==2), ]
NCbirths3<-NCbirths1[which(NCbirths1$Marital==1), ]
NCbirths4<-NCbirths1[which(NCbirths1$Marital==2), ]
```


This section of the analysis explores the relationship between the age of the subject and how much weight they gained after the pregnancy. Marital status is once again used as a categorical variable to see what differences there may be between married and unmarried subjects.


First, we look at a simple plot of all the observations, excluding cases with missing data:

```{r,fig.height = 3, fig.width = 3.8,echo=FALSE}
ggplot(NCbirths1, aes(MomAge, Gained, color=factor(marital_cat))) + geom_point(size=1) + labs(color="Marital Status")+geom_smooth(method = "lm") +ggtitle("NCbirths")

ggplot(NCbirths1, aes(MomAge, Gained, color=factor(marital_cat))) + geom_point(size=1) + labs(color="Marital Status")+geom_vline(xintercept = 27) + ggtitle("NCbirths1 with x-intercept of x = 27")

```

Compared to our original hypothesis of this relation, our inital plot shows little to no correlation between the two variables, with or without the categorical variable of marital status involved. However, when looking at the plots without any regression lines, a majority of observations that are unmarried occur roughly before the median of the age variable (27 years old). 

```{r, fig.height = 3, fig.width = 3.6,echo=FALSE}



ggplot(NCbirths2, aes(MomAge, Gained)) + geom_point(size=1) + labs(color="Marital") + geom_smooth(method = "lm") +ggtitle("Unmarried and 27 or Younger")

NCbirths1$marital_relevel <- factor(NCbirths1$Marital, levels=c("1","2"))
test1<-lm(Gained ~ MomAge*marital_relevel,data=NCbirths1)
#summary(test1)

#NCbirths2$marital_relevel <- factor(NCbirths2$Marital, levels=c("1","2"))
#test2<-lm(Gained ~ MomAge*marital_relevel,data=NCbirths2)
#summary(test2)
```

It appears that there could be a stronger relationship between age and weight gained if the dataset is simplified to just unmarried cases who are 27 or younger. To test this theory, we can check a comparison of the p-values of the age variable and R-squared values from the first plot to the second.


```{r,eval=FALSE,echo=FALSE}

fm<-lm(Gained~MomAge*marital_cat,data=NCbirths1)
fm1<-lm(Gained~MomAge, data = NCbirths2)
summary(fm)$coefficients
summary(fm1)$coefficients
summary(fm)$r.squared
summary(fm1)$r.squared
```


\ 

**Dataset    ** |  **p-value**     | **R-squared**
--------------  | ---------------- | -------- 
                |                  |
NCbirths1       |  0.249           |  0.0019
NCbirths2       |  0.258           |  0.0032

* NCbirths1 contains all observations that have no missing data
* NCbirths2 contains all observations of NCbirths1 that are unmarriend and 27 or    younger


After analysis, the new dataset has an increased p-value and is still above 0.05, so it tells us that age still has no statistical significance in relation to weight gained. However, the R-squared value did increase, albeit by an minimal amount. Perhaps this could lead to a greater relationship if the dataset contained more observations. But for now, with this dataset, it is evident that there was not strong enough of a relationship between age and weight gained to show a statistical significance between the two. This also goes against what our initial hypothesis stated, so we conclude that the original hypothesis was indeed false.

One last thing we can take away from the initial plot:

```{r,fig.height = 3, fig.width = 3.6,echo=FALSE}

ggplot(NCbirths3, aes(MomAge, Gained)) + geom_point(size=1) +geom_vline(xintercept = 27) + ggtitle("Married")

ggplot(NCbirths4, aes(MomAge, Gained)) + geom_point(size=1) +geom_vline(xintercept = 27) + ggtitle("Unmarried")



```

Once again, looking at the spread of observations in relation to the median age of 27, married women are distributed almost perfectly even from the youngest to oldest observations. But it's clear to see that, with unmarried women, after 27 the observations become very sparse. Before 27, there are many more observations and they're very condensed in that small section. Considering that the average age a women gets married in the U.S. is currently 27, this plot shows a strong confirmation of that statistic.


\newpage

# 3. Conclusions and Discussion

It should be noted that there are many missing values throughout the tables and analysis presented, which indicates an absence of data. For example, there is insufficient data on hawaiian individuals to apply a regression model for this specific subset. Additionally, there is a lack of data on premature birth data, which affected the accuracy of the analysis regarding weeks of gestation related to child weight. These missing values are easy easily overlooked in the midst of other discussion- but they indicate an underlying problem throughout the dataset. 

Additionally, it is easy and tempting to draw broad conclusions based on statistical analysis and tests like those applied earlier. However, it is important to remember that observing a trend does not allow mathematicians to draw conclusions about why or how those patterns emerge. For example, one might note that unmarried women tend to have children at a younger age than married women within the same racial subset- however, no conclusions should be drawn based on this observaton, especially at such initial stages of data analysis. Additionally, there is insufficient data to say that the observations made here are significant, or could apply to a larger population (the data here represents a convenience sample, and does not accurately depict the entire population). 

Despite these shortfalls, this data certainly tells an intriguing story. Interactions among variables related to the biology and social character of pregnant women provides unique insight into the complexity of childbirth, and the wide variety of factors that influence its unfolding. 
